---
title: "Areal Data Group Project"
author: "Kayla, Ingmar, Hanmo, Will"
date: "`r Sys.Date()`"
output: 
  html_document:
    keep_md: true
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Set up

**resources we are referencing:  

- http://www.css.cornell.edu/faculty/dgr2/_static/files/ov/ov_ADSA_Handout.pdf  
- Banerjee, Carlin and Gelfand, Hierarchical Modeling and Analysis for Spatial Data, 1st Edition, Ch 3  
- Bivand et al 2013 Ch 9  
- rspatial.org  

The packages we used are:  
```{r}
# install.packages(c("sp", "raster", "spdep", ....))
library(tidyverse)
library(sp)
# library(spData)
# library(raster) # or terra
library(terra)
library(spdep)
library(sf)
## more...
```

The dataset we are using is published in `spData` and comes from: _Anselin, Luc. 1988. Spatial econometrics: methods and models. Dordrecht: Kluwer Academic, Table 12.1 p. 189._
```{r}
## load dataset(s)
columbus <- vect(system.file("shapes/columbus.shp", package="spData")[1])
df.columbus <- as.data.frame(columbus)
glimpse(df.columbus)
```

The county data is:  

* HOVAL housing value (in \$1,000)  
* INC household income (in \$1,000)  
* CRIME residential burglaries and vehicle thefts per thousand households in the neighborhood  
* OPEN open space in neighborhood  
* PLUMB percentage housing units without plumbing  

Look at some summaries of those metrics: _Kayla_   
```{r, echo=FALSE, out.width= "50%"}
## set up plotting color palette
pal = colorRampPalette(c("lightgrey", "darkblue"))(10)

cc <- cut(columbus$HOVAL, 10)
mycols <- pal[as.numeric(cc)]
hist(df.columbus$HOVAL, 
     main = "Housing value", 
     xlab = "Value  in $1000")
plot(columbus, col = mycols, main = "Housing value in $1000")

cc <- cut(columbus$INC, 10)
mycols <- pal[as.numeric(cc)]
hist(df.columbus$INC, 
     main = "Household income", 
     xlab = "Value  in $1000")
plot(columbus, col = mycols, main = "Household income in $1000") 

cc <- cut(columbus$CRIME, 10)
mycols <- pal[as.numeric(cc)]
hist(df.columbus$CRIME,
     main = "Residential burglaries and vehicle thefts", 
     xlab = "CRIME per thousand households in the neighborhood")
plot(columbus, col = mycols, main = "Residential burglaries and vehicle thefts") 

cc <- cut(columbus$OPEN, 10)
mycols <- pal[as.numeric(cc)]
hist(df.columbus$OPEN, 
     main = "Open space in neighborhood", 
     xlab = "open space")
plot(columbus, col = mycols, main = "Open space in neighborhood") 

cc <- cut(columbus$PLUMB, 10)
mycols <- pal[as.numeric(cc)]
hist(df.columbus$PLUMB, 
     main = "Percentage of housing units without plumbing", 
     xlab = "Percent no plumbing")
plot(columbus, col = mycols, main = "Percentage of housing units without plumbing") 
```

Because we plan to use housing value as our response variable try a transformation.  
```{r}
hist(log10(df.columbus$HOVAL), 
     main = "Housing value", 
     xlab = "Value  in $1000 (log_10)")
```

# Exploratory Data Analysis  

## Measures of spatial association  

### Neighbors _Kayla_  
adjecency matrix construction - nearest neighbors, categories, 1/dist  
Figures out what the neighbors are:  
```{r}
xy <- centroids(columbus)
head(neighbors <- adjacent(columbus, symmetrical=TRUE))
plot(columbus, col='lightgray', border='black', lwd=1)
p1 <- xy[neighbors[,1], ]
p2 <- xy[neighbors[,2], ]
lines(p1, p2, col='red', lwd=2)
```

As an adjecency matrix  
```{r}
head(neighbors <- adjacent(columbus, pairs = FALSE))
```

### Moran's I - _Kayla_  

$$
I=\frac{n}{\sum_{i=1}^{n} (y_{i}-\overline{y})^{2}} \frac{\sum_{i=1}^{n} \sum_{j=1}^{n} w_{ij} (y_{i}-\overline{y}) (y_{j}-\overline{y})}{\sum_{i=1}^{n} \sum_{j=1}^{n} w_{ij}}
$$

Using `spdep` to test for spatial autocorrelation of the variable, by county in Columbus, OH. Moran's I is a global measure of spatial autocorrelation with values ranging -1 to 1. Here we are using a neighbor's matrix for any counties that are touching each other.   

The adjacency matrix:  
```{r}
ww <-  adjacent(columbus, "touches", pairs=FALSE)
```

Roughly the expected value for Moran's I is $E(I)=\frac{-1}{n-1}$   
```{r}
-1/(nrow(columbus)-1)
```

Values significantly ($\alpha = 0.05$) below that indicate negative spatial autocorrelation (a phenomena that generally occurs in random datasets) and above that indicates positive spatial autocorrelation (neighbors are more similar to each other than non-neighbors).  

*** I'm seeing the formula written in 2 ways above, and with denominator's flipped  

#### House value  
```{r, out.width= "50%"}
## Moran's I
(ac <- autocor(columbus$HOVAL, ww, "moran"))
## Monte Carlo sim to test for significance (I'm following https://rspatial.org/terra/analysis/3-spauto.html#compute-morans-i)
m <- sapply(1:99, function(i) {
    autocor(sample(columbus$HOVAL), ww, "moran")
})
hist(m)

## p-value
sum(m >= ac) / 100
```
```{r, echo=FALSE}
mi <- ac
mi.p <- ""
```
So there is not significant (Moran's I = `r ac`, p > 0.05) spatial autocorrelation.

#### Household income  
```{r, out.width= "50%"}
## Moran's I
(ac <- autocor(columbus$INC, ww, "moran"))
## Monte Carlo sim to test for significance (I'm following https://rspatial.org/terra/analysis/3-spauto.html#compute-morans-i)
m <- sapply(1:99, function(i) {
    autocor(sample(columbus$INC), ww, "moran")
})
# hist(m)

## p-value
sum(m >= ac) / 100
```
```{r, echo=FALSE}
mi[2] <- ac
mi.p[2] <- "*"
```
There is significant positive (Moran's I = `r ac`, p < 0.05) Spatial autocorrelation in household income.   

#### Crime  
```{r, out.width= "50%"}
## Moran's I
(ac <- autocor(columbus$CRIME, ww, "moran"))
## Monte Carlo sim to test for significance (I'm following https://rspatial.org/terra/analysis/3-spauto.html#compute-morans-i)
m <- sapply(1:99, function(i) {
    autocor(sample(columbus$CRIME), ww, "moran")
})
# hist(m)

## p-value
sum(m >= ac) / 100
```
```{r, echo=FALSE}
mi[3] <- ac
mi.p[3] <- "*"
```
So again we see significant spatial autocorrelation (Moran's I = `r ac`, p > 0.05).  

#### Open space  
```{r, out.width= "50%"}
## Moran's I
(ac <- autocor(columbus$OPEN, ww, "moran"))
## Monte Carlo sim to test for significance (I'm following https://rspatial.org/terra/analysis/3-spauto.html#compute-morans-i)
m <- sapply(1:99, function(i) {
    autocor(sample(columbus$HOVAL), ww, "moran")
})
# hist(m)

## p-value
sum(m >= ac) / 100
```
```{r, echo=FALSE}
mi[4] <- ac
mi.p[4] <- ""
```

There is not spatial autocorrelation with open space (Moran's I = `r ac`; p-value > 0.05)

#### Plumbing  
```{r, out.width= "50%"}
## Moran's I
(ac <- autocor(columbus$PLUMB, ww, "moran"))
## Monte Carlo sim to test for significance (I'm following https://rspatial.org/terra/analysis/3-spauto.html#compute-morans-i)
m <- sapply(1:99, function(i) {
    autocor(sample(columbus$HOVAL), ww, "moran")
})
# hist(m)

## p-value
sum(m >= ac) / 100
```
```{r, echo=FALSE}
mi[5] <- ac
mi.p[5] <- "*"
```

There is significant spatial autocorrelation with plumbing (Moran's I = `r ac`, p < 0.05).

### Geary's C  __Ingmar__

$$
C=\frac{(n-1) \sum_{i} \sum_{j} w_{i j}\left(Y_{i}-Y_{j}\right)^{2}}{2\left(\sum_{i \neq j} w_{i j}\right) \sum_{i}\left(Y_{i}-\bar{Y}\right)^{2}}
$$

Geary's C is a measure of local spatial autocorrelation that is roughly inversely related to Moran's I.  

#### House Value
```{r, out.width= "50%"}
(gearyc <- autocor(columbus$HOVAL, ww, "geary"))

## Monte Carlo sim to test for significance 
m <- sapply(1:99, function(i) {
    autocor(sample(columbus$HOVAL), ww, "geary")
})
hist(m)

## p-value
sum(m >= gearyc) / 100
```
```{r, echo = FALSE}
gc <- gearyc
```
No significant spatial autocorrelation (geary's c = `r gearyc`, p > 0.05).  

#### Household Income  
```{r}
(gearyc <- autocor(columbus$INC, ww, "geary"))

## Monte Carlo sim to test for significance 
m <- sapply(1:99, function(i) {
    autocor(sample(columbus$INC), ww, "geary")
})
# hist(m)

## p-value
sum(m >= gearyc) / 100
```
```{r, echo = FALSE}
gc[2] <- gearyc
```
No significant spatial autocorrelation (geary's c = `r gearyc`, p > 0.05).  

#### Crime  
```{r}
(gearyc <- autocor(columbus$CRIME, ww, "geary"))

## Monte Carlo sim to test for significance 
m <- sapply(1:99, function(i) {
    autocor(sample(columbus$CRIME), ww, "geary")
})
hist(m)

## p-value
sum(m >= gearyc) / 100
```
```{r, echo = FALSE}
gc[3] <- gearyc
```
No significant spatial autocorrelation (geary's c = `r gearyc`, p > 0.05).  

#### Open space  
```{r}
(gearyc <- autocor(columbus$OPEN, ww, "geary"))


## Monte Carlo sim to test for significance 
m <- sapply(1:99, function(i) {
    autocor(sample(columbus$OPEN), ww, "geary")
})
# hist(m)

## p-value
sum(m >= gearyc) / 100
```
```{r, echo = FALSE}
gc[4] <- gearyc
```
No significant spatial autocorrelation (geary's c = `r gearyc`, p > 0.05).  

#### Plumbing  
```{r}
(gearyc <- autocor(columbus$PLUMB, ww, "geary"))

## Monte Carlo sim to test for significance 
m <- sapply(1:99, function(i) {
    autocor(sample(columbus$PLUMB), ww, "geary")
})
# hist(m)

## p-value
sum(m >= gearyc) / 100
```
```{r, echo = FALSE}
gc[5] <- gearyc
```
No significant spatial autocorrelation (geary's c = `r gearyc`, p > 0.05).  

### Compare Moran's I and Geary's C  
Reinhard Furrer (http://user.math.uzh.ch/furrer/download/sta330/script_sta330.pdf, Version May 26, 2021) suggests to take 1-C to compare it to Moran's I more easily.  
```{r, echo=FALSE}
m <- data.frame(Morans_I = mi,
                signif = mi.p,
                Gearys_C = gc,
                One_minus_Gearys_C = 1-gc,
                signif = rep("", ),
                row.names = c("House value", "Income", "Crime", "Open space", "Plumbing"))
# m
knitr::kable(m, format = 'latex', digits = 3)
```

There are differences observed in local and global spatial autocorrelation in the data. For both Moran's I and Geary's C there was not significant spatial autocorrelation in house value or Open space. There was significant postive global spatial autocorrelation in household income, crime, and plumbing using Moran's I, but no significant local spatial autocorrelation.  

## Spatial smoothers  
Could delve further into this for the final project eg smoothing in image classification of fragmented habitats is problematic.  

### Clustering  

### Markov random fields  

# Spatial regression models

## Zero means _Ingmar_

$$
\mathit{HOVAL} = \frac{1}{n}\sum_{i=1}^{n}\mathit{HOVAL}
$$

```{r}
# make spatial vector to simple feature
columbus.sf <- sf::st_as_sf(columbus)
zero.means <- lm(HOVAL ~ 1, data=columbus.sf)
summary(zero.means)
```


## Simultaneous Autoregressive Models  

### SAR error model _Ingmar_

Use global moran's i to test model residuals for spatial correlation
lm.morantest()

```{r}
# make simple feature to neighborhood
columbus.nb <- poly2nb(columbus.sf)

# make neighborhood to list of weights
lw <- nb2listw(columbus.nb, style="W")

# estimate error SAR model without transformation
col.errW.eig <- errorsarlm(HOVAL~INC+CRIME+OPEN+CP, data=columbus.sf,
 lw, method="eigen", quiet=T)

# look at the residuals
hist(residuals(col.errW.eig))
```

```{r}
# estimate error SAR model with log transformation
col.errW.eig.log <- errorsarlm(log(HOVAL)~INC+CRIME+OPEN+CP, data=columbus.sf,
 lw, method="eigen", quiet=T)

hist(residuals(col.errW.eig.log))

# print model summary
summary(col.errW.eig.log, correlation=TRUE)
```


```{r}
# Run Moran's I test for col.errW.eig.log
moran.test(residuals(col.errW.eig.log), lw) # not significant
```


### SAR lag and Durbin models _Will_

### SAR lag and Durbin models _Will_

We compute the SAR lag and SAR Durbin models on the log-transformed housing value
with the same covariates used for the SAR error model in the previous section.

First we perform OLS regression and perform a (global?) Moran's I to test for
spatial dependence among the residuals.

(Procedure/code mirrors that in https://www.youtube.com/watch?v=b3HtV2Mhmvk)
```{r}
# define formula for log of HOVAL with covariates INC,CRIME,OPEN,CP
hoval.log <- log(HOVAL) ~ INC+CRIME+OPEN+CP

# ordinary least squares regression
regOLS = lm(hoval.log,data=columbus.sf)
summary(regOLS)
lm.morantest(regOLS,lw)
lm.LMtests(regOLS,lw,test="all")
```

The p-value of 0.002169 for the Moran's I statistic suggests spatial
dependence among residuals. Next we perform the SAR tests:

```{r}
library(rgdal)



# estimate SAR lag model with log transformation
lag.log <- lagsarlm(formula=hoval.log,data=columbus.sf,listw=lw,Durbin=FALSE)


hist(residuals(lag.log))

summary(lag.log, correlation=TRUE)
```

```{r}
# estimate SAR Durbin model with log transformation
durbin.log <- lagsarlm(formula=hoval.log,data=columbus.sf,listw=lw,Durbin=TRUE)

hist(residuals(durbin.log))
```

### Likelihood Ratio _Ingmar_

### Model comparisons  _Ingmar_

## Conditional Autoregressive Models  _Hanmo_

In the next step, we implement the conditional autoregressive (CAR) model on the Columbus data to study the impact of covariates on House values in Columbus, OH, 1980 with spatial information. 

The CAR model essentially assumes the spatial estimation is conditional on the value of neighbors. As a typical Bayes model, CAR model assumes prior distribution on the model parameters and applies computationally intensive sampling techniques like Markov Chain Monte Carlo (MCMC) or MCMC with Gibbs sampling to find the fitted parameters.



Since we have one response variable *House value*, the CAR model has only one random effect $\phi_k$ at each spatial location $k$. Because *House value* is continuous, Gaussian distribution is preferred for the CAR model and we take logarithm of the *House value* as well to make it normally distributed.


The package *CARBayes* by Duncan Lee is used to apply the CAR model in R. The specific function we use is *S.CARleroux*, where it specifies a CAR model proposed by Brian G. Leroux in 2000. The model expression is as follows. 

* $\mathbf{W}$ is a 0-1 neighborhood or weight matrix
* $k = 1, \dots, K$ is the index for a certain areal unit
* $\rho$ is a spatial correlation parameter, with $\rho=0$ means indenpendence and $\rho=1$ indicates strong spatial correlation


$$
\psi_{k}=\phi_{k}
$$
$$
\phi_{k} \mid \phi_{-k}, \mathbf{W}, \tau^{2}, \rho \sim \mathrm{N}\left(\frac{\rho \sum_{i=1}^{K} w_{k i} \phi_{i}}{\rho \sum_{i=1}^{K} w_{k i}+1-\rho}, \frac{\tau^{2}}{\rho \sum_{i=1}^{K} w_{k i}+1-\rho}\right)
$$

$$
\begin{aligned}
\tau^{2} & \sim \operatorname{Inverse}-\operatorname{Gamma}(a, b) \\
\rho & \sim \text { Uniform }(0,1)
\end{aligned}
$$


```{r echo=F,cache=T}
df.columbus_CAR = df.columbus
df.columbus_CAR$CP = as.factor(df.columbus_CAR$CP)
ggpairs(data = df.columbus_CAR, columns = c(7, 8:12, 18))
```


```{r cache=T}
set.seed(2021)
car_model_gaussian = CARBayes::S.CARleroux(log(HOVAL)~INC+CRIME+OPEN+DISCBD+CP, data = df.columbus_CAR, family = "gaussian", W = ww, burnin = 100000, n.sample = 300000, thin = 10, verbose = F)
```

```{r echo=F}
# print(paste0("The log likehood is ", round(logLik(car_model_gaussian), 2)))
# INC+CRIME+DISCBD+X+Y+PERIMETER+CP 13.59
```



```{r echo=F}
betas1 <- CARBayes::summarise.samples(car_model_gaussian$samples$beta, quantiles=c(0.5, 0.025, 0.975))
resultsCAR1 <- betas1$quantiles
rownames(resultsCAR1) <- colnames(car_model_gaussian$X)
knitr::kable(resultsCAR1, caption = "95% confidence intervals for model coefficients")

# print(paste0("The MSE is ", round(sqrt(sum((exp(car_model_gaussian$fitted.values) - df.columbus_CAR$HOVAL)^2)), 2)))

plot(x = df.columbus_CAR$HOVAL, y = exp(car_model_gaussian$fitted.values), type="p",
     xlab = "housing value", ylab = "Fitted value", main = "Fitted value vs Observed value")
abline(0,1)
```

The log likelihood of this model is 17.97  while the training root mean square error (RMSE) is 59.46. From the table of $95\%$ confidence intervals for coefficients, we have

* *CRIME* and *PLUMB* are the two variables that is significant since their confidence intervals don't have zero involved.
* Hold other predictors fixed, regions with **less crimes** tend to have higher *House values*.
* Hold other predictors fixed, regions with **higher household incomes** tend to have higher *House values*.
* Hold other predictors fixed, regions with **more open area** tend to have higher *House values*.
* Hold other predictors fixed, regions with **closer distance to CBD** tend to have higher *House values*.
* Hold other predictors fixed, **core** regions tend to have higher *House values*.

```{r echo=F}
cc <- cut(residuals(car_model_gaussian), 10)
mycols <- pal[as.numeric(cc)]
plot(columbus, col = mycols, main = "Residuals")
```

```{r}
# Moran I
Moran.I(residuals(car_model_gaussian), ww)$p.value
```

We also tested the Moran's I autocorrelation coefficient for the residuals. It shows there is **no** significant spatial correlation of the residuals since the p-value $p = 0.1 > 0.05$. Therefore, our CAR model fits the areal data nicely and leaves no significant spatial information in the residuals.
 

# Conclussions  

# References  

## Packages used  

```{r}
citation("tidyverse")
citation("sp")
citation("spData")
citation("terra")
citation("spdep")
```

## Other resources:  
```{r}
## rspataial, cornell thingy, 2 book chapters and paper Wendy shared
bibentry(
   bibtype = "Manual",
   title = "R Spatial",
   author = person("R Core Team"),
   organization = "R Foundation for Statistical Computing",
   address = "Vienna, Austria",
   year = 2013,
   url = "https://www.R-project.org/",
   key = "R")
```

